\label{sec:prob}


%\begin{center}
\section{Introduction}
%\end{center}

Distributed systems face a wide array of challenges such as \fixme{maintaining consistency, availability etc}.  While an ideal distributed system should tackle all of these challenges effectively, this is not feasible in practice. Each system is forced to make trade-offs on various aspects of the design. However, as distributed systems are designed to serve specific environments, we can optimize the system to tackle only those issues that are critical to the intended environment. Networks constitute one of the most popular and widely used distributed systems, and the networking domain presents us with a unique array of challenges. Our project attempts to solve the problem of traffic engineering in networks by extending existing ideas.

Routing decisions of shortest path routing protocols are solely based on pre-assigned costs in the network and are agnostic to the real traffic. This can lead to congestion in some parts of the network and under-utilization elsewhere. We can achieve greater responsiveness to traffic variations and freedom from hot-spots by relying on protocols that combine distance information with knowledge about congestion in the network. 

Backpressure routing protocol~\cite{BP-orig} is based on the idea that queue lengths provide direct indication of congestion in the network. Each node in the network decides the next hop for incoming traffic by comparing its queue length with that of its neighbors. The largest difference in queue length, i.e. the largest gradient, will occur along the best path towards destination. Thus, each node forwards traffic based on queue length information received from its immediate neighbors. 

Backpressure routing scheme is throughput optimal -- if the incoming traffic is within the capacity region of the network, the protocol will route it successfully. In spite of its throughput efficiency, backpressure protocol is not widely used in practice due to several limitations. First, the protocol does not consider path lengths. Forwarding decisions are solely based on local congestion information. This causes routing loops and delays in the network. Second, each node has to maintain separate queues for every destination in the network. This is impractical for conventional switches with limited buffer space. However, the protocol does provide interesting features such as throughput-optimal routing and congestion-awareness. 

Recently, several ideas have been proposed ~\cite{Srikant3, Austin1} that combine the notion of shortest path routing with backpressure protocol. This opens an exciting arena for congestion-aware shortest path routing. In particular, one of the variants~\cite{Srikant3} also obviates the need for multiple queues per node. But, the efficiency of the algorithm has not been tested extensively by the limited simulations performed on it. The algorithmic foundation also requires further work to be expanded into a fully-functional distributed protocol, suitable for practical realization. Our project aims at extending this work to develop a throughput-optimal congestion-aware distributed routing scheme. This involves optimizing the algorithm to speed up convergence and reduce delay in routing, with minimal communication overhead. We will design and deploy an optimized variant of the backpressure protocol in a testbed and analyze its performance with realistic traffic patterns.  

In the initial phase of our project, we developed a simulator to test the variations of the backpressure protocol and our proposed modifications. In this report, we present the results obtained on the simulator and a block diagram of our partially completed real-world implementation.